Higher-Order unification for free!

### 2024-07-15 paper/main.tex
minor rewording

  In this paper we propose to translate \Fo{} unification problems
  into an alternative encoding called \Ho, from which we derive \Uo,
  the higher-order unification procedure of \Fo{}. \Uo
  honours $\beta\eta$-equivalence for terms
  within the pattern fragment, and allows the use for 
  heuristics when unification is performed with terms outside the pattern fragment.
  Moreover, as \Uo{} delegates most of the work to \Ue,
  it can be used to efficiently simulate a logic program in \Fo{} by
  taking advantage of
  unification-related optimizations of the meta language, such as clause indexing.

@@

\begin{comment}
Specifying and implementing a proof system from scratch requires significant
effort. Logical Frameworks and Higher Order Logic Programming Languages provide
dedicated, high-level meta languages to facilitate this task in two ways: 1)
variable binding and substitution are for free when meta language binders
represent object logic ones; 2) proof construction, and
proof search, are greatly simplified by leveraging the unification procedure
provided by the meta language. Notable examples of meta languages are Elf,
Twelf, $\lambda$Prolog, Beluga, Abella and Isabelle which have been used to
implement or specify many formal systems such as First Order Logic, Set Theory,
Higher Order Logic, and the Calculus of Constructions.

The object logic we are interested in is Coq's type theory. We aim to develop a
higher-order unification-based proof search procedure using the meta language
Elpi, a dialect of $\lambda$Prolog. Elpi's equational theory includes
$\beta\eta$-equivalence and features a higher-order unification procedure \Ue
for the pattern fragment. Elpi offers an encoding of Coq terms that is suitable
for meta programming but that restricts \Ue to first-order unification problems
only. We refer to this basic encoding as \Fo.

In this paper we propose a more well-behaved encoding called \Ho, and show how
to translate unification problems from \Fo{} to corresponding ones in \Ho.
Consequently, we derive \Uo{}, the higher-order unification procedure of \Fo{}
that honours $\beta\eta$-equivalence (for Coq functions), addresses problems
within the pattern fragment, and allows for the use of heuristics to deal with
problems outside the pattern fragment. Moreover, as \Uo{} delegates most of the
work to \Ue, it can be used to efficiently simulate a logic program in \Fo{} by
taking advantage of unification-related optimizations of the meta language, such
as clause indexing.
\end{comment}

@@

The main difference is that we cannot implement \hrun since it is
the runtime of the programming language. In particular the runtime iterates
\Ue, but \hstep also needs to check links for \progress. Luckily Elpi extends
$\lambda$Prolog with constraints (suspended goals) and
Constraint Handling Rules (CHR) to operate on them~\cite{TASSI_2019,fruehwirth2017constraint}.
Similarly to \cite{pfenning1993}, a constraint is a delayed unification problems
suspended on a list of variables and is resumed \emph{as soon as one of these variable
is assigned}, before any other existing goal is considered. We point out that
we extend \cite{pfenning1993}, since we delay not only \notllambda unification problems, 
but also those problems that are still in \maybeeta.  
In turn link activation grants~\fullref{prop:fidelity}.
For brevity we only provide two pseaudo-code snippets. The first one depicts
how \linketa are suspended or make \progress.

>>>>

  In this paper we translate unification problems in \Fo{} 
  to an alternative encoding called \Ho, from which we derive \Uo,
  the higher-order unification procedure of \Fo{}. \Uo
  honours $\beta\eta$-equivalence for terms
  within the pattern fragment, and allows for the use of
  heuristics when the terms fall outside the pattern fragment.
  Moreover, as \Uo{} delegates most of the work to \Ue,
  it can be used to efficiently simulate a logic program in \Fo{} by
  taking advantage of
  unification-related optimizations of the meta language, such as clause indexing.

@@

@@

The main difference is that we cannot implement \hrun since it is
the runtime of the programming language. In particular the runtime iterates
\Ue, but \hstep also needs to check links for \progress. Luckily Elpi extends
$\lambda$Prolog with constraints (suspended goals) and
Constraint Handling Rules (CHR) to operate on them~\cite{TASSI_2019,fruehwirth2017constraint}.
Similarly to \cite{pfenning1993} a constraint is a goal % unification problems
suspended on a list of variables that is resumed \emph{as soon as one of these variable
is assigned}, before any other existing goal is considered. 
In turn link activation grants~\fullref{prop:fidelity}.
We point out that \cite{pfenning1993} delays \notllambda unification problems
while we also delay problems that are in \maybeeta.
For brevity we only provide two pseaudo-code snippets. The first one depicts
how \linketa are suspended or make \progress.
